---
title: "Les_Plan3_start"
author: "Charles Jason Tinant"
date: "1/27/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We usually think about precipitation as a **depth** rather than a volume.  We can use the equation $Vol = depth * area$.  To find out the average **depth** of water that falls on each country each year, we need to divide the land area from the precipitation volume. However, in the example below, we don't have the total land area of the world.  However, from Table 1.1 in the text we have the land areas by continent. 

```{r calculate_ann_depth_km}
# we can quickly add data together in R rather than read it in from a
# .csv file using the 'concatinate' function, c().  Than using the 
# 'tidyverse', we can easily bind rows and columns together.

library(tidyverse) # load the tidyverse package

# make a vector of strings with continent names and a vector of area values 
continent <- c("Africa", "North America", "South America", "Asia", 
               "Europe", "Oceania")
area_thos_km2 <- c(3022, 2471, 1784, 4458, 1018, 853)

# lets turn these into data.frames rather than vectors
continent <- as.data.frame(continent)
area_thos_km2 <- as.data.frame(area_thos_km2) 

# bind the two vectors together and then make it into a data-frame
land_area_km2 <- bind_cols(continent, area_thos_km2)

# Take a look at land_area to check if it looks ok. Now use 'sum()' and
# use the result to find the global average annual precipitation depth.
# Note: the *1000 at the end is to get the land area into kilometers
land_area_tot_km2 <- sum(land_area_km2$area_thos_km2) *1000

precip_ann_km <- precip_ann_km3 / land_area_tot_km2
```

Of course, we don't express the depth in kilometers.  We can use the following converstion factor  *We use a lot of unit conversion in hydrology.*  

$depth(m) = depth(km) *1000(m/km)$

```{r calculate_ann_depth_m}
# use a conversion factor to find the average global precipitation 
# depth in meters

precip_ann_m <-precip_ann_km * 1000 # 1000 m in 1 km
print(precip_ann_m)
```


**Get into groups of about four and write your own equation to find the *effective precipitation*, the amount of precipitation after evaporation**


