---
title: "NSci233 Notes- Week 4"
author: "Charles Jason Tinant"
date: "2/06/2018"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: Lesson Plan 3 - Precipitation
fontsize: 12pt
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Updates following Feb 6 lecture
* We will be having a supplementary / make-up class this Friday from 1-4 PM to work with R-Studio and the rest of Lesson Plan 2.  Note this is a rechedule of the last supplementary / make-up class that had to be cancelled due to weather.

## Overview
Precipitation is the single most important variable driving hydrologic processes and events, because it replenishes soil moisture, lakes and streams, and groundwater.  Precipitation is any form of solid or liquid water the falls from the atmosphere to the earth's surface.  Precipitation can be thought about in terms of events, as averages over a period of time, or as deviations from the normal or "adequate" amount.  In this section we are going to discuss the role of precipition in climate, what causes precipiation, precipitation events, and rainfall frequency. 

##Student Learning Outcomes (SLOs) &  Lesson Goals:  

- *SLO1: Students will demonstrate proficiency in the components of the hydrologic cycle: precipitation, infiltration, runoff, and evapotranspiration* - Students will be able to generally describe what causes precipiation, precipitation events, and rainfall frequency. 

- *SLO2: Students are able to analyze components of the hydrologic with computer software* - Students will be generally able to read  data into R, and perform some analysis using packages.  Students will will be able to read arguments of a package from R Documentation.

## Readings
- Read sections 2.7 - 2.8 of the text 
- Article on Capetown water supply

## Homework
#You are welcome to turn in a file in Rmd, pdf, html or Word.  Upload the file into a folder with your name in Google Drive. 

#1. Vocabulary - Summarize the terms: weather, climate, 

2. [Koeppen Climate](http://koeppen-geiger.vu-wien.ac.at/pdf/KG_USA_UScounty.pdf) - What is the climate for Western South Dakota?  What is the climate for Eastern South Dakota.

2. [Drought Monitor](http://droughtmonitor.unl.edu/) - What is the drought intensity for this week for the county you live in? If your county is in drought, then is the drought short-term or long-term?  What is the drought summary for the High Plains?  What is the the drought summary for the West?

#3. Water Distribution - Is ground water a renewable resource? Use Table 1.8 to identify how long it takes for a drop of water removed from an aquifer to be replaced.

#4.  

## Lecture Notes 

### Weather, Climate, and Water Resources
**Weather** is the result of the day-to-day movement of moist air masses from thier sources and thier associated effects on precipitation and temperature.  **Climate** is a long-term average of the weather.  Intermediate to weather and climate is a moving window of recent weather history and intermediate term forecasting, and how these relate to the climate.  One tool that describes the weather trends and relates them to projections is the Drought Monitor (http://droughtmonitor.unl.edu/).  

A lot of schemes have been used to describe climate.  One of the most widely adopted schemes is the Koeppen Climate Classification System.  Check out the changing climate at http://koeppen-geiger.vu-wien.ac.at/. 

Köppen’s classification is based on a subdivision of terrestrial climates into five major types, which are represented by the capital letters A, B, C, D, and E. Each of these climate types except for B is defined by temperature criteria. Type B designates climates in which the controlling factor on vegetation is dryness (rather than coldness). Aridity is not a matter of precipitation alone but is defined by the relationship between the precipitation input to the soil in which the plants grow and the evaporative losses. Since evaporation is difficult to evaluate and is not a conventional measurement at meteorological stations, Köppen was forced to substitute a formula that identifies aridity in terms of a temperature-precipitation index (that is, evaporation is assumed to be controlled by temperature). Dry climates are divided into arid (BW) and semiarid (BS) subtypes, and each may be differentiated further by adding a third code, h for warm and k for cold.

Temperature defines the other four major climate types. These are subdivided, with additional letters again used to designate the various subtypes. Type A climates (the warmest) are differentiated on the basis of the seasonality of precipitation: Af (no dry season), Am (short dry season), or Aw (winter dry season). Type E climates (the coldest) are conventionally separated into tundra (ET) and snow/ice climates (EF). The mid-latitude C and D climates are given a second letter, f (no dry season), w (winter dry), or s (summer dry), and a third symbol (a, b, c, or d [the last subclass exists only for D climates]), indicating the warmth of the summer or the coldness of the winter. Although Köppen’s classification did not consider the uniqueness of highland climate regions, the highland climate category, or H climate, is sometimes added to climate classification systems to account for elevations above 1,500 metres (about 4,900 feet). (source: https://www.britannica.com/science/Koppen-climate-classification)
  

** Get into a group of 3-4 and use Table 2.1 the Koeppen climate map at http://koeppen-geiger.vu-wien.ac.at/pdf/KG_USA_UScounty.pdf to identify the climate of South Dakota.  Why is ranching the predominant land use west of the Missouri and farming the predominant land use east of the Missouri?**

## Causes of Precipitation


**Get into a group of about four.  Run the code below and discuss: 1. the purpose what '(10^9)/(10^6)' is doing, 2. the overall meaning of the water_withdraw dataset using the table in terms of wealth and diet, and 3. the overall meaning of the water_withdraw dataset using the graphic in terms of wealth and diet**

```{r join_population}
# We can add to our data table using 'join'. 
# We are also using a 'pipe operator' %>% to make reading 
# the code a little bit easier.

# First, bring in a table of regional use
regional_use <- read_csv("regional_use.csv") %>%
  select(-X1)   # this gets rid of a useless column

# Next, use a 'full_join' to join the tables together by similiar regions.
water_withdraw <- full_join(water_withdraw, regional_use, by = "region")

rm(regional_use)
# If these steps were performed correctly, you should have the first 
# six rows of Table 1.3.  We can find out how much water is being
# being used per capita using the 'mutate' function to create a new column

water_withdraw <- water_withdraw %>%
  mutate(withdraw_m3_pers_yr  = withdraw_bill_m3_yr / pop_mill * 
                           ((10^9)/(10^6))) %>%
  arrange(desc(withdraw_bill_m3_yr))

# Sometimes percentages are difficult relate.  This changes the percentages
#   to how much water per person is going into different water uses
water_withdraw <- water_withdraw %>%
  mutate(ag_m3_pers_yr = withdraw_m3_pers_yr * agric_perc/100) %>%
  mutate(ind_m3_pers_yr = withdraw_m3_pers_yr * indust_perc/100) %>%
  mutate(munic_m3_pers_yr = withdraw_m3_pers_yr * munic_perc/100) %>%
  arrange(desc(ag_m3_pers_yr)) 

# Sometimes, a graphic helps to to see what is happening in a data-set.
#  1. Gather the data to graph it easier.
water_withdraw2 <- water_withdraw %>%
  select(-(2:7)) %>%
  gather(key = type, val = vol_m3, -region)

# 2. Create the graph.
ggplot(water_withdraw2, aes(x = region, y = vol_m3, color = type)) +
  geom_jitter() +
  labs(title = "Per capita water withdrawals by use and region",
       subtitle = "2010 data") +
  ylab("cubic meters") +
  xlab("") + 
  theme_bw() +
  theme(legend.position = "bottom")
```

The graph above is why they say in the United States that "whiskey is for drinking and water is for fighting over...". In North America, about equal amounts of water are withdrawn for industry and agriculture. Also, note North America uses approximately twice the world average for municipal use. 

*Question: what water resource is being used during drought?*

The graph above shows water withdrawal, which is different from water consumption.   A majority of industrial and municipal water is discharge as wastewater; however, a substantial portion of agricultural water is consumed.  And much of the water used for agriculture is ground water, which is not a renewable resource.

So, how does this information work together to help us make water resource decisions?  One take away from the exercise above is that you need to have data in a format that allows you to look at the data across categories.  Above, we normalized the data by per-capita.  Below, we are going to normalize the data by area.  Consider that flow rate is volume per time $Q = [m^3/year]$  Volume is made up of depth times area $V = [m^3] = d[m]*A[m^2]$.  So, to compare water volumes across regions of different size, we can divide the volume of water per year for a region by the area of the region to find the depth.  

**Get into a group of about four.  Run the code below and discuss: 1. the purpose what '(10^9)/(10^6)' is doing, 2. the overall meaning of the water_withdraw dataset using the table in terms of wealth and diet, and 3. the overall meaning of the water_withdraw dataset using the graphic in terms of wealth and diet**
```{r water_balance}
water_balance <- water_withdraw %>%
  select(region, withdraw_bill_m3_yr) %>%
  rename(withdraw_bill_m3 = withdraw_bill_m3_yr)

balance_tabl <- read.csv("water_balance.csv") %>%
  select(-X)

water_balance <- full_join(water_balance, balance_tabl)

# bill = 10^9
# mill = 10^6
# km^2 = 10^6 m
# So, for the equation withdraw_bill_m3 / area_mill_km^2, the
# units are 10^9 m3 / 10^12 m3 -> 1/10^3 m3
# so, using 1 m = 10^2 cm, we can get to cm if we divide the equation by 10

water_balance <- water_balance %>%
  mutate(withdraw_cm = withdraw_bill_m3 / (10 * area_mill_km2)) %>%
  mutate(withdraw_cm = round(withdraw_cm, digits = 2)) %>%
  arrange(desc(withdraw_cm)) %>%
  mutate(withdraw_precip_rat = withdraw_cm/precip_cm) %>%
  mutate(withdraw_precip_rat = round(withdraw_precip_rat, digits = 2))

# Sometimes, a graphic helps to to see what is happening in a data-set.
#  1. Gather the data to graph it easier.
water_balance2 <- water_balance %>%
  select(-withdraw_bill_m3, -area_mill_km2, -withdraw_precip_rat) %>%
  rename(pcp = precip_cm) %>%
  rename(gw = gw_runoff_cm) %>%
  rename(sw = sw_runoff_cm) %>%
  rename(evap = evaporation_cm) %>%
  rename(use = withdraw_cm) %>%
  gather(key = type, val = depth_cm, -region) %>%
  mutate(type = as.factor(type))

#  1. set up print order of the factors 
water_balance2$type <- factor(water_balance2$type, 
                              levels = c("pcp", "evap", "sw", "gw", "use"))

# 2. Create the graph.
ggplot(water_balance2, aes(x = type, y = depth_cm, color = type)) +
  geom_point() +
  facet_wrap(~region) +
  labs(title = "Water balance by world region",
       subtitle = "2010 data") +
  ylab("depth in cm") +
  xlab("") + 
  theme_bw() +
  theme(legend.position = "bottom")
```




```{r background_work, eval=FALSE, include=FALSE}
library(janitor)
world_pop <- read.csv("world_pop.csv")
pop_2010 <- world_pop %>%
  clean_names() %>%
  rename(pop_mill = x2010) %>%
  select(region, pop_mill)

# add data for Table 1.3 in text
region <- c("Africa", "North America", "South America", "Asia", "Europe", "Oceania")
region <- as.data.frame(region)

agric_perc <- c(82, 44, 67, 80, 22, 60)
agric_perc <- as.data.frame(agric_perc)

indust_perc <- c(5,42, 11, 10, 57, 15)
indust_perc <- as.data.frame(indust_perc)

munic_perc <- c(13, 15, 22, 10, 22, 26)
munic_perc <- as.data.frame(munic_perc)

use_tabl <- bind_cols(region, agric_perc, indust_perc, munic_perc)
pop_2010 <- read.csv("pop_2010.csv")

use_tabl <- full_join(use_tabl, pop_2010)

write.csv(use_tabl, "regional_use.csv")

# add data for Table 1.5 in text
region <- c("Europe", "Asia", "Africa", "North America", "South America", "Oceania")
region <- as.data.frame(region)

area_mill_mi2 <- c(3.8, 17.6, 11.8, 8.1, 4.0, 3.4)
area_mill_km2 <- area_mill_mi2*(5/3.2)
area_mill_km2 <- round(as.data.frame(area_mill_km2), digits = 2)

precip <- c(28.9, 28.6, 27.0, 26.4, 64.9, 29)
precip <- as.data.frame(precip)

gw_runoff <- c(4.3, 3.0, 1.9, 3.3, 8.3, 2.1)
gw_runoff <- as.data.frame(gw_runoff)

sw_runoff <- c(8.3, 8.5, 3.6, 8.0, 14.7, 2.8)
sw_runoff <- as.data.frame(sw_runoff)

evaporation <- c(16.3, 17.0, 21.5, 15.1, 41.9, 20.1)
evaporation <- as.data.frame(evaporation)

balance_tabl <- bind_cols(region, precip, gw_runoff, sw_runoff, evaporation)

balance_tabl <- balance_tabl %>%
  mutate_all(funs(cm = . / 2.54)) %>%
  select(-region_cm) %>%
  mutate_if(is.numeric, funs(round(., digits = 2))) %>%
  select(-(2:5))

balance_tabl <- bind_cols(area_mill_km2, balance_tabl)
balance_depth <- balance_tabl %>%
  mutate_if(is.numeric, funs(. * area_mill_km2))

write.csv(balance_tabl, "water_balance.csv")
```


